---
title: "Plos_One_paper_analysis"
author: "Hannah Comiskey"
date: "2025-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```


## Introduction


This vignette demonstrates how to reproduce the analysis from: “Bayesian probabilistic projections of proportions with limited data: An application to subnational contraceptive method supply shares.”

The workflow uses the mcmsector package to:
  1. Prepare and clean subnational family planning data
  2. Construct spline bases and indexing needed for the model
  3. Fit a hierarchical Bayesian JAGS model
  4. Extract posterior samples of contraceptive method proportions
  5. Summarise and visualise posterior estimates against the observed data

Each section explains what the code is doing and how it contributes to the final output.

## Load and explore the subnational data

This section loads required libraries and prints a glimpse of the raw dataset included in the package. The subnat_bivar_data dataset contains subnational contraceptive method share data for multiple countries, sectors, and time periods.

```{r subnational data}
library(dplyr)
library(rjags)
library(R2jags)
library(tidyr)
library(ggplot2)
library(mcmsector)

glimpse(subnat_bivar_data)
```

## Prepare data

Before fitting the model, the raw survey input must be cleaned, harmonised across years, and mapped into the structure required by JAGS.
What this step does:
  1. Removes unusable or incomplete survey observations
  2. Applies DEFT adjustments to account for survey design
  3. Creates indices for: Country, subnational region contraceptive method and time period
  4. Constructs B-spline basis matrices (B_ik) for each subnational region
  6. Generates metadata used throughout the modelling workflow

These steps are performed by two convenience functions: `clean_fp_source_data()` and `run_preprocessing_pipeline()`

```{r prepare data, echo=FALSE}

data <- clean_fp_source_data(mcmsector::subnat_bivar_data, deft_lookup = mcmsector::DEFT_DHS_database)

inputs <- run_preprocessing_pipeline(raw_df = data, 
                                     area_classification = mcmsector::Country_and_area_classification_inclFP2020,
                                     methods = c("Female Sterilization", "Implants", "Injectables", "IUD", "OC Pills"),
                                     all_years = seq(1990, 2030.5, by = 0.5),
                                     nseg = 10)


```

## Transform observed proportions to the logit scale

The JAGS model operates on logit-transformed proportions, as these lie on the real line and avoid boundary issues at 0 and 1. This section computes the logit of the public sector proportions and corresponding standard errors via the delta method.

```{r run JAGS model, echo=FALSE}
mydata <- inputs$data[,c("Public", "Public.SE")]
logit.data <- mydata %>%
  rowwise() %>%
  mutate(logit.Public = log(Public/(1-Public)),
         logit.Public.Var = ((1/(Public*(1-Public)))^2)*Public.SE^2,
         logit.Public.SE = sqrt(logit.Public.Var))
n_method = c("Female Sterilization", "Implants", "Injectables", "IUD", "OC Pills")
M =  length(n_method)

```


## Build JAGS model input list

This step assembles all required components into a single list (inputdata) that JAGS can read. Components include:
  1. Observed logit proportions and SEs
  2. B-spline basis matrices
  3. Dimensions of the model (methods, regions, years)
  4. Matching indices linking observations to hierarchical levels
  5. A prior covariance matrix `Omega` for method-level effects

```{r}
## The required data ------------------------------
inputdata <- list(y = as.vector(unlist(logit.data[,c("logit.Public")])), # using total proportions as collapsing over sectors
                  se_prop = as.vector(unlist(logit.data[,c("logit.Public.SE")])),
                  Bik = inputs$splines$B_ik,
                  n_years = inputs$meta$n_years,
                  n_obs = nrow(logit.data),
                  K = inputs$splines$K,
                  H = inputs$splines$H,
                  kstar = inputs$splines$Kstar,
                  C_count = length(inputs$meta$n_country),
                  P_count = length(inputs$meta$n_subnat),
                  M_count = length(c("Female Sterilization", "Implants", "Injectables", "IUD", "OC Pills")),
                  Omega = diag(M) * 0.01 + diag(M),
                  matchcountry= inputs$meta$index_country_subnat_tbl$index_country,
                  matchsubnat = inputs$data$index_subnat,
                  matchmethod = inputs$data$index_method,
                  matchyears = inputs$data$index_year)
```

## Fit the Bayesian JAGS model

We run the multivariate non-centered parameterisation (NCP) model with smooth splines (ncp_MVN_SEmodel_JAGS.txt). This model produces posterior samples for:
  1. alpha_pms: method × subnational random effects
  2. beta.k: spline coefficients
  3. Covariance matrices for method-country and method-subnational random effects
  4. A residual variance parameter sigma_delta

```{r}
pars <- c( "alpha_pms", # required for P
          "inv.Sigma.alpha_cms",
          "inv.Sigma.alpha_pms",
          "sigma_delta",
           "beta.k")

model_path <- system.file("model", "ncp_MVN_SEmodel_JAGS.txt", package = "mcmsector")

mod <- jags.parallel(data=inputdata,
            parameters.to.save=pars,
            model.file = model_path,
            n.iter = 70000,         # total number of iterations per chain
            n.burnin = 10000,
            n.thin=30)

saveRDS(mod, file='JAGS_model_MVN_NCP_kstar_SE.RDS')

```

## Set up indexing tables for summarisation

These tables help convert JAGS indices back into meaningful labels (method, region, year, sector) when summarising posterior draws.

```{r}
method_index_table <- tibble(index_method = 1:length(n_method), Method = n_method)
sector_index_table <- tibble(index_sector = 1:2, Sector = c('Public', 'Private'))

all_years = seq(1990, 2030.5, by = 0.5)

year_index_table <- tibble(average_year = all_years, index_year = 1:length(all_years))
subnat_index_table <- inputs$data %>% select(Country, Region, index_country, index_subnat) %>% distinct()

```

## Extract posterior parameters

We extract the `alpha_pms` and `beta_k` samples and package them into a format ready for transformation back to proportions.

```{r, eval=FALSE}
params <- extract_mcmsector_params(mod = readRDS('JAGS_model_MVN_NCP_kstar_SE.RDS'), 
                                   n_method = length(n_method),
                                   n_subnat = length(inputs$meta$n_subnat))
```

## Simulate posterior proportions

We reconstruct posterior samples of the public/private method proportions using:
  1. Intercepts from alpha_pms
  2. Spline predictions from beta_k
  3. Logistic transformation back to the probability scale

```{r, eval=FALSE}
prop_samps <- simulate_posterior_proportions(alpha_pms = params$alpha_pms,
                                           beta_k = params$beta_k,
                                           B_ik = inputs$splines$B_ik,
                                           n_method = length(n_method),
                                           n_subnat = length(inputs$meta$n_subnat),
                                           all_years = seq(1990, 2030.5, by = 0.5),
                                           n_samps = 4000)
```

## Summarise posterior draws

We compute a tidy long-format table merged with indexing definitions and:
  1. posterior means
  2. 95% credible intervals

This object can then be used for plotting and comparison with observed survey data.

```{r, eval=FALSE}
prop_summary <- summarise_posterior_proportions(P = prop_samps,
                                                method_index_table = method_index_table,
                                                sector_index_table = sector_index_table,
                                                subnat_index_table = subnat_index_table,
                                                year_index_table = year_index_table)

```

## Visualise the model estimates

Below we illustrate how to reproduce the plots used in the paper. This example filters to Rwanda and four methods for clarity.
The plot overlays the observed survey shares with 95% uncertainty, posterior mean trajectories and posterior predictive credible intervals.

```{r, eval=FALSE}

P_df<- inputs$data  %>%
  select(Country, Region, Method, index_year, average_year, Public, Private) %>%
  left_join(method_index_table) %>%
  pivot_longer(cols = c(Public, Private), names_to = 'Sector', values_to = 'Observed')

P_SEdf<- inputs$data  %>%
  select(Country, Region, Method, index_year, average_year, Public.SE, Private.SE) %>%
  left_join(method_index_table) %>%
  pivot_longer(cols = c(Public.SE, Private.SE), names_to = 'Sector', values_to = 'SE') %>%
  mutate(Sector = stringr::str_replace(Sector, '.SE', ''))

P_df <- left_join(P_df, P_SEdf) %>%
  rowwise() %>%
  mutate(lower_95 = Observed - 2*SE,
         upper_95 = Observed + 2*SE)


rwanda_obs <- P_df %>% filter(Country=="Rwanda" & Method %in% c("Implants", "Injectables", "IUD", "OC Pills")) 
rwanda_preds <- prop_summary %>% filter(Country=="Rwanda" & Method %in% c("Implants", "Injectables", "IUD", "OC Pills")  & average_year < 2027)

ggplot() +
  geom_point(data = rwanda_obs, aes(x=average_year, y=Observed, colour=Sector, pch=Sector)) +
  geom_errorbar(data = rwanda_obs, aes(x=average_year, ymin=lower_95, ymax=upper_95, colour=Sector)) +
  geom_line(data = rwanda_preds, aes(x=average_year, y=Mean, colour=Sector, lty=Sector)) +
  geom_ribbon(data = rwanda_preds, aes(x=average_year, ymin=lower_95, ymax = upper_95, fill=Sector), alpha=0.2) +
  theme_minimal() +
  theme(title = element_text(size=20), axis.text.x = element_text(angle = 90), strip.text.x = element_text(size=20), strip.text.y = element_text(size=20), axis.title.x = element_text(size=20), axis.title.y = element_text(size=20)) +
  theme(legend.position = "bottom", legend.title = element_text(size = 20), legend.text = element_text(size = 20))+
  facet_grid(Region ~ Method)


```






